---
title: "Increased Importance of Altruistic Factors in Jobs"
subtitle: "Analysis of US General Social Survey from 1989 to 2016"
author: 
- Shirley Chen
- Jessica Im
thanks: "Code and data in this report are available at: https://github.com/shirleychen003/job_satisfaction.git."
date: today
date-format: "long" 
format: pdf
abstract: Rising altruism has been reported globally; however, many altruistic occupations have continued to be undervalued in society. To investigate perceived importance of altruistic motivations in work, data from the U.S. General Social Survey on the reported importance of jobs that help others and have social usefulness from 1989 to 2016 was analyzed. Our exploration revealed that the importance of altruistic motivations has increased overall across 35 years, that women view them as more important than men but the gap has decreased over time, and that older participants tend to view the motivations as more important and most between ages 35 and 44. These findings highlight the various factors influencing perspectives on the importance of altruistic occupations, and underscore a need for further research on impacts of rising altruism in occupational perspectives on care work and desired careers.
editor: visual
toc: true
number-sections: true
bibliography: bibliography.bib
---

```{r}
#| include: false
### Workspace Setup ###
library(tidyverse)
library(tinytex)
library(janitor)
library(here)
library(knitr)
library(usethis)
library(gitcreds)
library(kableExtra)

#### Import data ####
# Question 1 Dataset
cleaned_GSS_data <-
  read_csv("../data/cleaned_GSS_data.csv")

cleaned_GSS_data_renamed <-
  read_csv("../data/cleaned_GSS_data_renamed.csv")

#### Variables ####

ordered_responses <- c('very_important', 
                       'important',
                       'neither',
                       'not_important',
                       'not_important_at_all')
legend_labels <- c('Very Important', 
                   'Important',
                   'Neither',
                   'Not Important',
                   'Not Important At All')

custom_colours <- c('#FDE726', 
                   '#440154',
                   '#3B528B',
                   '#21908D',
                   '#5DC863')

chosen_responses <- c("very_important", "important")

total_participants <-cleaned_GSS_data_renamed |>
  summarise(count = n())

```

# Introduction
Altruism has been on the rise and in a 2016 report by the Charities Aid Foundation, volunteering, charitable giving, and providing everyday help had increased for the previous five years [@marsh]. Altruism is defined by Merriam-Webster as the unselfish quality possessed by those who focus on the welfare of others [@merriamwebster]. While altruism is generally associated with donating and selfless acts of kindness, how altruistic motivations are impacted in the context of job satisfaction has not been thoroughly explored. Altruistic occupations tend to have harsher consequences on a worker’s mental and physical health; however, many altruistic jobs are still undervalued in society. The COVID-19 pandemic worsened pre-existing strains on healthcare labour shortages and the overworked, underpaid nurses took the brunt of the damage. In 34 out of 49 countries assessed by the International Labour Organization, despite their extensive contributions to a global pandemic, nurses and midwives were paid below average wage for high-skilled workers [@ILO2023]. To understand why care workers are undervalued and what changing perspectives may exist, perspectives on the importance of altruistic work must be investigated.

Interested in American perspectives on the importance of a job that helps others and a job that is useful to society, we gathered data from the United States General Social Surveys (GSS) [@GSS]. We analyzed our estimand, the proportion of responses in favour of our two intrinsically motivating factors being important in 1989, 1998, 2006, and 2016. Additionally, comparing this data to the ages and gender of participants allowed further investigation of what factors could influence responses. By investigating the perceived importance of these altruistic motivators, a more comprehensive perspective on what influences the occupations Americans seek and prioritize is revealed.

Overall, reported importance of the two factors increased from 1989 to 2016 in importance. We also found that female respondents felt the factors were more important than the male respondents did, but the difference decreased from 1989 to 2016. Finally, analyzing the relationship between age and responses revealed...!!!

In [Data], the data collection methodology, the data cleaning processes, and key variables. In [Results], the relationship between demographic information of participants and their responses as well as the overall change in importance is analyzed with supportive figures. Finally, in [Discussion], the significance and possible causes of our findings are explored within economic stability, gender, age, and culture. 


# Data

## Source Data and Methodology

Based at the University of Chicago since 1972, the GSS is a project with the objective of monitoring and analyzing the intricacies of American society [@GSSabout]. The GSS Data Explorer makes it so that data retrieved from the project is a publicly available resource, accessible to various types of people, such as educators, policymakers, or researchers through the National Opinion Research Center (NORC). It has even been referenced in leading publications, such as the New York Times and the Associated Press. Due to the variety of subjects covered regarding American society, the GSS is one of the most frequently visited resources for information for social sciences. 

Majority of the GSS data was collected through face-to-face interviews with the target population of adults (18+) residing in the United States. The standard national survey methods were practiced, such as hiring interviewers and training supervisors when needed. Interviewers were required to complete a practice interview supervised by evaluators at NORC [@GSScodebook]. However, starting in 2002, computer-assisted personal interviewing (CAPI) methods were introduced [@GSScodebook]. The use of manual edits and keypunching were eliminated, and training to learn how to use CAPI was included. 

The dataset used for this paper was retrieved from the GSS Data Explorer website [@GSSexplorer]. All the survey data used to measure job satisfaction was in relation to job and work in the Work Orientation Module during the years 1989, 1998, 2006, and 2016; the specific variable names extracted from the dataset being intjob, hlpoths, and hlpsoc. For the years and demographic data, the specific variable names extracted were year, age and sex. 



## Data Cleaning

The open source statistical programming language [@R-base] was used to clean and analyze the data, along with producing the graphs. The main packages that supported this process included [@R-tidyverse], [@R-ggplot2], [@R-knitr], [@R-janitor], and [@R-lubridate].

The cleaning process involved filtering the specific data variables used for our analysis from the downloaded GSS dataset, and renaming any variables with meaningful names. For example, rather than “hlpoths” being the column name for “importance of helping others in a job” , we renamed it to helping_others, as shown in Table #. Further, the numerical values representing the participants’ responses (1-5) were changed to the representative words/phrases (not important, very important, etc.). @tbl-cleaneddata shows the old and new variable names used in cleaning, the description of variables, and sample responses. 

```{r}
#| label: tbl-cleaneddata
#| echo: false
#| warning: false
#| tbl-cap: GSS Dataset

variable_data <- data.frame( 
  Variable = c("hlpoths","hlpsoc"), 
  "New Name" = c("helping_others", "social_usefulness"), 
  Description = c("Importance of helping others in a job",
                  "Importance of social usefulness in a job"), 
  "Example Response" = c("Very Important", "Not Important")
)

kable(variable_data, format = "simple")
```

## Data Terminology

The response choices for each question and their respective code in brackets are as follows: Inapplicable (-100), No Answer (-99), Do Not Know/Cannot Choose (-98), Very Important (1), Important (2), Neither (3), Not important (4), and Not Important At All (5). For our graphs, we did not include the Inapplicable, No Answer, and Do Not Know/Cannot choose responses to focus on the discernible participant responses.

## Respondent Demographics
@tbl-gender shows the number and percentage of male and female respondents for 1989, 1998, 2006, and 2016. The percentages of female participants were consistently higher than the male participants, as the female participant percentages were always above 50% while the male participant percentages ranged from low to high 40s.

@tbl-agegroups displays the number of respondents within different age groups for 1989, 1998, 2006, and 2016. The classified age groups are ‘18-24’, ‘25-34’, ‘45-54’, ‘55-64’, and ‘65+’. Further, there is a column labelled “N/A”  for the respondents who did not disclose their age. As shown by the table, the 18-24 age group had the least amount of participants every year, while the age group with the most participants per year varied; however, the age range of the most participants per year stayed between 25-54.

```{r}
#| label: tbl-gender
#| echo: false
#| warning: false
#| tbl-cap: Respondent Gender Demographics for 1989, 1989, 2006, and 2016

gender_demographic_data <- cleaned_GSS_data |>
  filter(year %in% c(1989, 1998, 2006, 2016)) |>
  group_by(year, sex) |>
  summarise(count = n()) |>
  mutate(percentage = round((count / sum(count)) * 100, 2))

kable(gender_demographic_data, col.names = c("Year","Sex", "Count","Percentage"),
      format = "simple")
```


```{r}
#| label: tbl-agegroups
#| echo: false
#| warning: false
#| tbl-cap: Respondent Count of Participants in Age Groups by Year

age_group_1989 <- cleaned_GSS_data |>
  filter(year == 1989) |>
  mutate(age_group = case_when(
    age >= 18 & age <= 24 ~ "18-24",
    age >= 25 & age <= 34 ~ "25-34",
    age >= 35 & age <= 44 ~ "35-44",
    age >= 45 & age <= 54 ~ "45-54",
    age >= 55 & age <= 64 ~ "55-64",
    age >= 65 ~ "65+"
  )) |>
  count(age_group) |>
  pivot_wider(names_from = age_group, values_from = n) |>
  rename("N/A" = `NA`)

age_group_1998 <- cleaned_GSS_data |>
  filter(year == 1998) |>
  mutate(age_group = case_when(
    age >= 18 & age <= 24 ~ "18-24",
    age >= 25 & age <= 34 ~ "25-34",
    age >= 35 & age <= 44 ~ "35-44",
    age >= 45 & age <= 54 ~ "45-54",
    age >= 55 & age <= 64 ~ "55-64",
    age >= 65 ~ "65+"
  )) |>
  count(age_group) |>
  pivot_wider(names_from = age_group, values_from = n) |>
  rename("N/A" = `NA`)

age_group_2006 <- cleaned_GSS_data |>
  filter(year == 2006) |>
  mutate(age_group = case_when(
    age >= 18 & age <= 24 ~ "18-24",
    age >= 25 & age <= 34 ~ "25-34",
    age >= 35 & age <= 44 ~ "35-44",
    age >= 45 & age <= 54 ~ "45-54",
    age >= 55 & age <= 64 ~ "55-64",
    age >= 65 ~ "65+"
  )) |>
  count(age_group) |>
  pivot_wider(names_from = age_group, values_from = n) |>
  rename("N/A" = `NA`)

age_group_2016 <- cleaned_GSS_data |>
  filter(year == 2016) |>
  mutate(age_group = case_when(
    age >= 18 & age <= 24 ~ "18-24",
    age >= 25 & age <= 34 ~ "25-34",
    age >= 35 & age <= 44 ~ "35-44",
    age >= 45 & age <= 54 ~ "45-54",
    age >= 55 & age <= 64 ~ "55-64",
    age >= 65 ~ "65+"
  )) |>
  count(age_group) |>
  pivot_wider(names_from = age_group, values_from = n) |>
  rename("N/A" = `NA`)

# Put the Year column back
age_group_1989 <- mutate(age_group_1989, Year = 1989)
age_group_1998 <- mutate(age_group_1998, Year = 1998)
age_group_2006 <- mutate(age_group_2006, Year = 2006)
age_group_2016 <- mutate(age_group_2016, Year = 2016)

combined_age_group_table <- bind_rows(age_group_1989,
                                      age_group_1998,
                                      age_group_2006,
                                      age_group_2016)

combined_age_group_table <- combined_age_group_table |>
  select(Year, everything())

kable(combined_age_group_table, format = "simple")
  
```


```{r}
#| label: tbl-age
#| echo: false
#| warning: false
#| tbl-cap: Respondent Age Demographics by Year

age_1989_data <- cleaned_GSS_data |>
  filter(year==1989) |>
  select('age') |>
   summarise(
    mean = round(mean(age, na.rm = TRUE)),
    median = round(median(age, na.rm = TRUE)),
    mode = as.numeric(names(sort(table(age), decreasing = TRUE))[1]),
    min = min(age, na.rm = TRUE),
    max = round(max(age, na.rm = TRUE))
  )

age_1998_data <- cleaned_GSS_data |>
  filter(year==1998) |>
  select('age') |>
   summarise(
    mean = round(mean(age, na.rm = TRUE)),
    median = round(median(age, na.rm = TRUE)),
    mode = as.numeric(names(sort(table(age), decreasing = TRUE))[1]),
    min = min(age, na.rm = TRUE),
    max = round(max(age, na.rm = TRUE))
  )
age_2006_data <- cleaned_GSS_data |>
  filter(year==2006) |>
  select('age') |>
   summarise(
    mean = round(mean(age, na.rm = TRUE)),
    median = round(median(age, na.rm = TRUE)),
    mode = as.numeric(names(sort(table(age), decreasing = TRUE))[1]),
    min = min(age, na.rm = TRUE),
    max = round(max(age, na.rm = TRUE))
  )

age_2016_data <- cleaned_GSS_data |>
  filter(year==2016) |>
  select('age') |>
   summarise(
    mean = round(mean(age, na.rm = TRUE)),
    median = round(median(age, na.rm = TRUE)),
    mode = as.numeric(names(sort(table(age), decreasing = TRUE))[1]),
    min = min(age, na.rm = TRUE),
    max = round(max(age, na.rm = TRUE))
  )

years <- c("1989", "1998", "2006", "2016")
combined_age_data <- bind_rows(age_1989_data, 
                               age_1998_data, 
                               age_2006_data, 
                               age_2016_data)
combined_age_data <- cbind(years, combined_age_data)


kable(combined_age_data, col.names = c("Year",
                                       "Mean", 
                                       "Median",
                                       "Mode", 
                                       "Min", 
                                       "Max"), format = "simple")
```

## Graphs of Responses
@fig-helpsothers and @fig-useful, shows the responses to the prompt "On the following list there are various aspects of jobs. Please circle one number to show how important you personally consider it is in a job" where each graph represents one of the aspects. Respondents answered on a 1 to 5 Likert scale where 1 represents "very important" and 5 represents "not important at all". 

### Helps Others
In @fig-helpsothers, the proportion of respondents to the prompt "A job that allows someone to help other people?" is displayed. From the first year of data collection in 1989 to 2006, "Important" was the most selected response. In 2016, "Very Important" surpassed "Important" by 1%. In general, you can see an increase in "Very Important" respondents across the years while there is little change in the proportion of "Not Important" and "Not Important At All" responses. Further, there is a general decrease in “Neither” responses from 1989 - 2006 which is interrupted when there is a slight increase in 2016.


```{r}
#| label: fig-helpsothers
#| echo: false
#| warning: false
#| fig-cap: Q1 - "A job that allows someone to help other people?"

# Helping Others Data
helping_others_data <- cleaned_GSS_data_renamed[c('year',
                                          'helping_others')]

helping_others_percentages <- helping_others_data |>
  count(year, helping_others) |>
  group_by(year) |>
  mutate(percentage = n / sum(n) * 100) 

helping_others_percentages$helping_others <- factor(
  helping_others_percentages$helping_others, levels =
    ordered_responses)

ggplot(helping_others_percentages, aes(
  x = as.factor(year),
  y = percentage, 
  fill = helping_others)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Helping Others Responses by Year",
       x = "Year",
       y = "Percentage",
       fill = "Responses") +
  scale_y_continuous(labels = scales::percent_format(scale = 1), 
                     limits = c(0, 100)) + 
  scale_fill_manual(values = custom_colours,
                    breaks = c('very_important', 
                       'important',
                       'neither',
                       'not_important',
                       'not_important_at_all'),
                    labels = c('Very Important', 
                   'Important',
                   'Neither',
                   'Not Important',
                   'Not Important At All')) + 
  theme_minimal()
```


### Social Usefulness
@fig-useful displays the proportion of responses for the prompt "A job that is useful to society?". There is a large increase in the proportion of "Very Important" responses from 1989 to 2016. In contrast, there is a gradual decline for both "Important" and "Neither". There is little change in "Not Important" and "Not Important At All". Compared to the other figures, this graph has the most varying change in the “neutral” response.

```{r}
#| label: fig-useful
#| echo: false
#| warning: false
#| fig-cap: Q3 - "A job that is useful to society?"

# Social Usefulness Data
social_usefulness_data <- cleaned_GSS_data_renamed[c('year', 'social_usefulness')]

social_usefulness_percentages <- social_usefulness_data |>
  count(year, social_usefulness) |>
  group_by(year) |>
  mutate(percentage = n / sum(n) * 100) 

social_usefulness_percentages$social_usefulness <- factor(
  social_usefulness_percentages$social_usefulness, levels = ordered_responses)

ggplot(social_usefulness_percentages, aes(
  x = as.factor(year),
  y = percentage, 
  fill = social_usefulness)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Social Usefulness Responses by Year",
       x = "Year",
       y = "Percentage",
       fill = "Responses") +
  scale_y_continuous(labels = scales::percent_format(scale = 1), 
                     limits = c(0, 100)) + 
  scale_fill_manual(values = custom_colours,
                    breaks = c('very_important', 
                       'important',
                       'neither',
                       'not_important',
                       'not_important_at_all'),
                    labels = c('Very Important', 
                   'Important',
                   'Neither',
                   'Not Important',
                   'Not Important At All')) + 
  theme_minimal()
```


# Results

## Overall Trends

@tbl-avgresponses summarizes the average responses per year for each question, where 1 represents “Very Important” and 5 represents “Not Important At All”. There is a general decrease in scores, meaning importance increased, each year. The largest change from 1989 to 2016 was “Social Usefulness”, with an average response decrease of 0.311. Overall, interesting work was consistently more important than helping others and social usefulness with an average of 0.264 points lower. As time passed, interesting work became rated more similarly with the other two factors, and in 2016 the average difference was only 0.144 points. 

```{r}
#| label: tbl-avgresponses
#| tbl-cap: Average of Responses by Year
#| warning: false
#| echo: false

helping_others_avg <- cleaned_GSS_data |>
  group_by(year) |>
  summarize(
    avg_helping_others = round(mean(helping_others), 3))

social_usefulness_avg <- cleaned_GSS_data |>
  group_by(year) |>
  summarize(
    avg_helping_others = round(mean(social_usefulness), 3))

combined_table <- cbind(
  helping_others_avg,
  social_usefulness_avg[, -1])

kable(combined_table,
      format = "markdown", 
      col.names = c("Year",
                    "Helping Others",
                    "Social Usefulness"),
      digits = 3,
      booktabs = TRUE,
      linesep = "")
```

### Gender

In @tbl-helpingothers1989 and @tbl-helpingothers2016, we see the percentage distribution of responses by respondent sex for the importance of helping others. In 1989, female respondents took the more extremes, and had much higher percentages for “very important” compared to their male counterparts. In 2016, the trend continues but at a lesser scale, with the proportions evening more.

```{r}
#| label: tbl-helpingothers1989
#| tbl-cap: Helping Others Response Proportions 1989
#| warning: false
#| echo: false

HO_demographic_1989 <- 
  cleaned_GSS_data_renamed[c('year',
                             'helping_others',
                             'sex')] 

HO_demographic_1989 <- HO_demographic_1989 |>
  filter(year==1989) |>
  mutate(response = factor(helping_others, levels = ordered_responses)) |>
  group_by(response, sex) |>
  summarise(count = n()) |>
  mutate(percentage = round((count / sum(count)) * 100, 2))

HO_demographic_1989$response <- fct_recode(HO_demographic_1989$response,
                                           "very important" = "very_important",
                                           "important" = "important",
                                           "neither" = "neither",
                                           "not important" = "not_important",
                                           "not important at all" = "not_important_at_all")

kable(HO_demographic_1989, col.names = c('Response',
                                         'Sex',
                                         'Count',
                                         'Percentage'), format = "simple")
```

```{r}
#| label: tbl-helpingothers2016
#| tbl-cap: Helping Others Response Proportions 2016
#| warning: false
#| echo: false

HO_demographic_2016 <- 
  cleaned_GSS_data_renamed[c('year',
                             'helping_others',
                             'sex')] 

HO_demographic_2016 <- HO_demographic_2016 |>
  filter(year==2016) |>
  mutate(response = factor(helping_others, levels = ordered_responses)) |>
  group_by(response, sex) |>
  summarise(count = n()) |>
  mutate(percentage = round((count / sum(count)) * 100, 2))

HO_demographic_2016$response <- fct_recode(HO_demographic_2016$response,
                                           "very important" = "very_important",
                                           "important" = "important",
                                           "neither" = "neither",
                                           "not important" = "not_important",
                                           "not important at all" = "not_important_at_all")

kable(HO_demographic_2016, col.names = c('Response',
                                         'Sex',
                                         'Count',
                                         'Percentage'), format = "simple")
```


In @tbl-socialusefulness1989 and @tbl-socialusefulness2016, the previous trend continues, with more women favouring importance and a decrease in gender differences in 2016. Social usefulness and interesting work had similar differences in proportions, while 1989 helping others had the largest disparity.

```{r}
#| label: tbl-socialusefulness1989
#| tbl-cap: Social Usefulness Response Proportions 1989
#| warning: false
#| echo: false

SU_demographic_1989 <- 
  cleaned_GSS_data_renamed[c('year',
                             'social_usefulness',
                             'sex')] 

SU_demographic_1989 <- SU_demographic_1989 |>
  filter(year==1989) |>
  mutate(response = factor(social_usefulness, levels = ordered_responses)) |>
  group_by(response, sex) |>
  summarise(count = n()) |>
  mutate(percentage = round((count / sum(count)) * 100, 2))

SU_demographic_1989$response <- fct_recode(SU_demographic_1989$response,
                                           "very important" = "very_important",
                                           "important" = "important",
                                           "neither" = "neither",
                                           "not important" = "not_important",
                                           "not important at all" = "not_important_at_all")

kable(SU_demographic_1989, col.names = c('Response',
                                         'Sex',
                                         'Count',
                                         'Percentage'), format = "simple")
```

```{r}
#| label: tbl-socialusefulness2016
#| tbl-cap: Social Usefulness Response Proportions 2016
#| warning: false
#| echo: false

SU_demographic_2016 <- 
  cleaned_GSS_data_renamed[c('year',
                             'social_usefulness',
                             'sex')] 

SU_demographic_2016 <- SU_demographic_2016 |>
  filter(year==2016) |>
  mutate(response = factor(social_usefulness, levels = ordered_responses)) |>
  group_by(response, sex) |>
  summarise(count = n()) |>
  mutate(percentage = round((count / sum(count)) * 100, 2))

SU_demographic_2016$response <- fct_recode(SU_demographic_2016$response,
                                           "very important" = "very_important",
                                           "important" = "important",
                                           "neither" = "neither",
                                           "not important" = "not_important",
                                           "not important at all" = "not_important_at_all")

kable(SU_demographic_2016, col.names = c('Response',
                                         'Sex',
                                         'Count',
                                         'Percentage'), format = "simple")

```

### Age

@tbl-age_helpingothers

```{r}
#| label: tbl-age_helpingothers
#| tbl-cap: Helping Others Response Percentages (%) by Age
#| warning: false
#| echo: false
#| message: false

age_group_HO <- cleaned_GSS_data_renamed |>
  group_by(helping_others) |>
  filter(helping_others %in% chosen_responses) |>
  mutate(age_group_HO = case_when(
    age >= 18 & age <= 24 ~ "18-24",
    age >= 25 & age <= 34 ~ "25-34",
    age >= 35 & age <= 44 ~ "35-44",
    age >= 45 & age <= 54 ~ "45-54",
    age >= 55 & age <= 64 ~ "55-64",
    age >= 65 ~ "65+"
  )) |>
  count(age_group_HO) |>
  pivot_wider(names_from = age_group_HO, values_from = n) |>
  rename("Response" = `helping_others`) |>
  mutate(Response = fct_recode(Response, "very important" = "very_important"))

age_group_HO <- age_group_HO |>
  select(-`NA`)

age_group_percentage_HO <- age_group_HO |>
  mutate(`18-24` = round(`18-24` / total_participants$count * 100, 2),
         `25-34` = round(`25-34` / total_participants$count * 100, 2),
         `35-44` = round(`35-44` / total_participants$count * 100, 2),
         `45-54` = round(`45-54` / total_participants$count * 100, 2),
         `55-64` = round(`55-64` / total_participants$count * 100, 2),
         `65+` = round(`65+` / total_participants$count * 100, 2))

kable(age_group_percentage_HO, format = "simple")

```

@tbl-age_socialusefulness

```{r}
#| label: tbl-age_socialusefulness
#| tbl-cap: Social Usefulness Response Percentages (%) by Age
#| warning: false
#| echo: false
#| message: false

age_group_SU <- cleaned_GSS_data_renamed |>
  group_by(social_usefulness) |>
  filter(social_usefulness %in% chosen_responses) |>
  mutate(age_group_SU = case_when(
    age >= 18 & age <= 24 ~ "18-24",
    age >= 25 & age <= 34 ~ "25-34",
    age >= 35 & age <= 44 ~ "35-44",
    age >= 45 & age <= 54 ~ "45-54",
    age >= 55 & age <= 64 ~ "55-64",
    age >= 65 ~ "65+"
  )) |>
  count(age_group_SU) |>
  pivot_wider(names_from = age_group_SU, values_from = n) |>
  rename("Response" = `social_usefulness`) |>
  mutate(Response = fct_recode(Response, "very important" = "very_important"))

age_group_SU <- age_group_SU |>
  select(-`NA`)

age_group_percentage_SU <- age_group_SU |>
  mutate(`18-24` = round(`18-24` / total_participants$count * 100, 2),
         `25-34` = round(`25-34` / total_participants$count * 100, 2),
         `35-44` = round(`35-44` / total_participants$count * 100, 2),
         `45-54` = round(`45-54` / total_participants$count * 100, 2),
         `55-64` = round(`55-64` / total_participants$count * 100, 2),
         `65+` = round(`65+` / total_participants$count * 100, 2))

kable(age_group_percentage_SU, format = "simple")
```


## Change in importance over time
@fig-allhelpingothers demonstrates the changing proportion of responses to how important is it for a job to help others. 

```{r}
#| label: fig-allhelpingothers
#| echo: false
#| message: false
#| warning: false
#| fig-cap: Proportion of Responses to Importance of Job that Helps Other

# Prepare data
helping_others_data <- cleaned_GSS_data[c('year',
                                          'helping_others')]

helping_others_data <-helping_others_data |>
  mutate(
    helping_others = recode(helping_others, 
                            '1' = 'Very Important',
                            '2' = 'Important',
                            '3' = 'Neither',
                            '4' = 'Not Important',
                            '5' = 'Not Important At All'),
  )

helping_others_data <-
  helping_others_data |>
  mutate(
    year = (as.character(year)),
  )

helping_others_percentages <- 
  helping_others_data |>
  count(year, helping_others) |>
  group_by(year) |>
  mutate(percentage = n / sum(n) * 100)


# Graph data
helping_others_percentages |>
  ggplot(mapping = aes(x = year, y = percentage, fill = helping_others)) +
  geom_col() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 10)) +
  labs(x = "Year", y = "Proportion for each response (%)", fill = "Response") +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(size = 5)) +
  scale_fill_viridis_d(option = "viridis") 
```


```{r}
#| label: fig-allsocialusefulness
#| echo: false
#| message: false
#| warning: false
#| fig-cap: Proportion of Responses to Importance of Social Usefulness in Job

# Prepare data
social_usefulness_data <- cleaned_GSS_data[c('year',
                                          'social_usefulness')]

social_usefulness_data <-
  social_usefulness_data |>
  mutate(
    social_usefulness = recode(social_usefulness, 
                            '1' = 'Very Important',
                            '2' = 'Important',
                            '3' = 'Neither',
                            '4' = 'Not Important',
                            '5' = 'Not Important At All'),
  )

social_usefulness_data <-
  social_usefulness_data |>
  mutate(
    year = (as.character(year)),
  )

social_usefulness_percentages <- 
  social_usefulness_data |>
  count(year, social_usefulness) |>
  group_by(year) |>
  mutate(percentage = n / sum(n) * 100)


# Graph data
social_usefulness_percentages |>
  ggplot(mapping = aes(x = year, y = percentage, fill = social_usefulness)) +
  geom_col() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 10)) +
  labs(x = "Year", y = "Proportion for each response (%)", fill = "Response") +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(size = 5)) +
  scale_fill_viridis_d(option = "viridis") 
  scale_fill_viridis_d(option = "viridis") 
```


# Discussion

## Economic Factors
In the early 1990s, much of the Western world, including the United States, entered a recession [@wikiemployment1990s]. The 1980s had a very slow productivity growth at 1.5 each year and at the end of the early 90s recession, the Federal budget had a $298 billion deficit [@useconomy]. In 2000, the U.S. had their highest unemployment rate in three decades at 4.0% [@useconomy]. These low employment rates from late 1989 to 1992 likely contributed to the lower amount of importance workers placed on their occupation being one which helps others and is useful to society; instead, workers may have prioritized pay and stability. In contrast, the 2010s had the lowest unemployment rate, low inflation rates, and low interest [@pbs2010seconomy]. The decade was the first since the 1850s to skip a recession and low unemployment broke a 50-year record. These factors may have contributed to participants putting higher value on altruistic occupation factors.


## Gender
Women generally placed more importance on occupations that help others and are useful to society. The disparity between attitudes of women and men was largest for the importance of occupations that help others, and it is evident how this translates to women working in care work. In 2021, 77.6% of the 21.2 million U.S. care workers were women [@womencareworkus]. Care work includes occupations in which workers care for others like children, persons with disabilities, seniors, and include industries such as education, healthcare, personal services, and more [@womencareworkcanada]. Despite the essential work the care industry does, it has historically been underpaid work and home health care workers had an average salary of $13.81 from 2018 to 2020 [@womencareworkusepi]. The issue worsens when intersectional impacts are considered, as care workers are disproportionately women of colour. The change in attitude disparity between 1989 and 2016 may be a sign of more supportive views for care work and lowered inequality; however, the continued low pay of care workers is evidence that changing attitudes have not largely improved the care economy. The general trend of lower importance across both factors for men may also be an indication of persisting stereotypes that men need to prioritize pay and supporting their own family. This may be a feedback loop, where men believe they must be making more money to support their family while women believe their value is in care work, and when men receive higher pay than their female counterparts it is confirmed that they are responsible for monetarily supporting the family. 


## Age


## Culture
Generally, the data shows an increasing support for occupations that help others and contribute to society. These factors can be categorized into intrinsic motivations which are aspects of an occupation which are desired because they are enjoyable in themselves, while extrinsic motivations refer to aspects sought for reasons outside of work [@diffintrinsicextrinsic]. In individualistic cultures, intrinsic motivations have a stronger link to job satisfaction whereas in collectivist countries, extrinsic factors are intrinsically motivating [@monnot], [@hu]. [@santos] has found that there are more individualistic relational practices in the United States and that the use of individualistic words has increased over time. The relation between individualistic culture and intrinsic motivations in combination with increasing individualism in the U.S. may explain the increase in importance of occupations with intrinsic value. The U.S. also scores below average on uncertainty avoidance according to the Culture Factor Group, perhaps contributing to an increased openness to taking a riskier job that may be more rewarding [@hofstede]. An exploration to pursue in the future is comparing various intrinsically and extrinsically motivating factors with changes in individualism and uncertainty avoidance over time and across cultures. 

# References

# Appendix

## Survey
Survey available at: https://forms.gle/ezXF9ADu7phXJGM87

### Survey Introduction
The following survey aims to investigate the importance of various factors in jobs. From analyzing 1989 to 2016 data in the United States General Social Survey, our paper revealed the increasing importance of intrinsic motivations over 35 years. To deepen our understanding of the relationship between demographic factors and the importance of occupational motivations, we created this short survey. 

### Questions
1. All responses to this survey are completely confidential and will only be used for research purposes. Your email will only be used if you request your responses to be sent to you. You may withdraw consent and close this survey at any time. For any further inquiries, please email jessica.im@mail.utoronto.ca or sshirleyy.chen@mail.utoronto.ca.
* I consent.

2. What is your age group?
  + 18-24
  + 25-34
  + 35-44
  + 45-54
  + 55-64
  + 65+

3. What gender do you identify with?
  + Woman
  + Man
  + Non-Binary
  + Other:

4. What race do you identify with?
  + Asian or Pacific Islander
  + Black or African American
  + Hispanic or Latino
  + First Nations or Indigenous
  + White or Caucasian
  + Multiracial/Biracial
  + Other:

5. What nationality/nationalities do you identify with?
  + Type...

6. What is your current employment status?
  + Full-time employee
  + Part-time employee
  + Unemployed
  + Retired
  + Student
  + Other:

For the following questions, rank them on the how important you personally consider it is in a job. 
7. What is the importance of a job that allows someone to help other people?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)
8. What is the importance of a job that pays well?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)
9. What is the importance of a job that is interesting?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)
10. What is the importance of job security?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)
11. What is the importance of a job that is collaborative?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)
12. What is the importance of a job that has autonomy?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)
13. What is the importance of a job that is useful to society?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)
14. What is the importance of a job that has growth opportunities?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)
15. What is the importance of a job that has work-life balance?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)
16. What is the importance of a job that has flexible hours?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)
17. What is the importance of a job that has a flexible location?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)
18. What is the importance of a job that has a positive social impact?
  + Likert scale 1 (Very Important) - 5 (Not Important At All)

### Survey Confirmation Message
Thank you for completing our survey, your submission has been recorded and your answers may be requested to be sent to the email you entered. Your participation is valued and impactful in our quest to better understand occupational motivations. By participating, you have helped reveal what types of jobs are desired and valued.
